\large How resource abundance and stochasticity affect animals' spatial needs

### 2023-07-27 writing/manuscript.Rmd
updated based on Sarah's comments (2)

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Simon Wood
    institute: uofe
  # - name: Adam T. Ford
  #   institute: biol
  #   institute: braes
  # - name: Jacob R. Goheen
  #   institute: wyom
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
  # - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

Here, we illustrate how animals' spatial needs depend on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource *stochasticity* on animals spatial needs while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resource abundance is often unpredictable (and hard to quantify), since it depends on a variety of factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differents in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time. Quantifying resources as a numerical random variable (as opposed to using *ad hoc* qualitative descriptions) provides us with the capacity to leverage techniques from probability theory and statistics.

@@

\noindent In this section, we use empirical tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. We estimate $\e(R)$ and $\var(R)$ by modeling NDVI [@pettorelli_normalized_2011] using a Beta location-scale GAM, after scaling NDVI to [0, 1]. Appendix 2 contains additional information on how NDVI was modeled for the empirical example, along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.

The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., figure \@ref(fig:5-5-sims)) and generalize results to a single, common set of functional responses. We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all motile organisms, once differences in size, metabolic needs, and diet are accounted for. In this subsection, we illustrate how researchers can leverage these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings. Additionally, the analyses can be expanded to the population or species level with the use of Hierarchical Generalized Additive Models with individual-level intercepts and slopes [see models "GS" and "GI" in @pedersen_hierarchical_2019]. Such models would allow one to test whether members of the same population or species behave according to common functional responses to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior. In the following sections, we illustrate how one can apply this framework to empirical data.

We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs. Next, we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$. Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.

@@

\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e. its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary. `mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs. Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie families.

The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative. However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g. available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and NDVI [@pettorelli_normalized_2011].

The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day). The Tweedie location-scale family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density. In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family (not available in `mgcv` at the time of publication). If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].

Irrespectively of whether one uses a frequentist or Bayesian approach, it is important to model the mean and variance terms carefully. Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and non-monotonic, we suggest using a GAM rather than a GLM. However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the location terms. An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated. Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt is a reasonable starting point. Additionally, using REstricted Marginal Likelihood [REML, a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results. We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals. Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies and with a large sample size (but low effective sample size due to the spatiotemporal autocorrelation).

@@

\noindent In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011]. However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the **linear transformation**

@@

\noindent where $\nu$ indicates the NDVI value [see @denny_fallacy_2017 for more information on transforming response data]. <!--Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

@@

\noindent As discussed in the previous section, the first step to modeling $H$ is to choose an appropriate distribution. Since $H$ is strictly positive and continuous, a Gamma distribution seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017]. The Tweedie distribution may also be appropriate, if the mean-variance relationship of the Gamma distribution results in a poor fit.

As with the simulated effects of $\mu(t)$ and $\sigma^2(t)$ on $H$ (figure \@ref(fig:5-5-reg)), a location-scale Gamma model is likely required. If there is not sufficient data to fit a location scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best family of distributions using an information criterion such as the Akaike Information Criterion [@akaike_new_1974]. While a GAM would allow the effects of $\e(R)$ and $\var(R)$ on $H$ to be quite flexible, we suggest one use a GLM, instead, (unlike when modeling $R$) because we expect the relationship to be fairly simple and monotonic. Restricting the model to a GLM reduces the risk of issues that may arise due to limited sample size and temporal autocorrelation in $\e(R)$ and $\var(R)$. If one is interested in fitting a GAM but imposing shape constraints, such as monotonicity, we suggest fitting a shape-constrained GAM via the `scam` package [as we do here: see the section below, Appendix 2, and @pya_scam_2022].

@@

```{r tapir-mw, fig.cap="Seven-day home-range size of a tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance. (a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir. (b.) Varince in resource abundance, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir. (c.) Estimated seven-day home range based on the 95\\% utilization quantiles. (d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size. The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them. (f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated. Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$. See Appendix 2 for additional information. The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
knitr::include_graphics('../figures/tapir-example.png')
```

\noindent Figure \@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$. Panels **a** and **b** show the changes in mean and variance in NDVI, respectively, during the tracking period. Panel **c** shows the changes in the tapir's space use over time. Note how the home-range size is larger during periods of lower NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018). In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable. Finally, panels **d** and **e** show the estimated (partial) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use. Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)). Thus, while the time series shown in panel (c.) may appear complicated, the tapir's response can be well described by two simple functional responses.

@@

\noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs. The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings. Instead, one can test whether individuals follow these hypotheses, if members of the same population or species behave similarly, and if all members of the group have a common functional response [using hierarchical models, see @pedersen_hierarchical_2019]. Individuals and populations that were thought to have different behaviors and spatial needs due to differences in behavior may instead be simply responding to different local conditions along the same functional response [e.g., @singh_migration_2012].

@@

\noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events which occur at the scale of the organism [i.e. approximately one or more times per generation and on the spatial scale of the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants, *Loxodonta sp.*: @foley_severe_2008; muskoxen, *Ovibos moschatus*: @berger_climate_2018; bison, *Bison bison*, and mule deer, *Odocoileus hemionus*: @geremia_migrating_2019]. Consequently, organisms with long generations (e.g., elephants) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus respond to and prepare for stochastic events by adjusting the timing of reproduction [framewokrk: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hybernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006], moving long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017] including migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], and developing more complex cognitive abilities over longer lifespans, generation times, and developmental periods [@brown_toward_2004], including the development of memory of such events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021]. However, adapting to changes in environmental cues rather than relying on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].

>>>>

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Simon Wood
    institute: uofe
  # - name: Adam T. Ford
  #   institute: biol
  #   institute: braes
  # - name: Jacob R. Goheen
  #   institute: wyom
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
  # - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    -  \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

Here, we illustrate how animals' spatial needs depend on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource *stochasticity* on animals' spatial needs while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resource abundance is often unpredictable (and hard to quantify), since it depends on a variety of factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differents in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

@@

\noindent In this section, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals via remote sensing data such as NDVI [@pettorelli_normalized_2011]. Appendix 2 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.

The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., figure \@ref(fig:5-5-sims)) and generalize results to a single, common set of functions. We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all motile organisms, once differences in size, metabolic needs, and diet are accounted for. In this subsection, we illustrate how researchers can leverage these hypotheses using their own data to build on previous work rather than listing the results as new findings. Additionally, the analyses can be expanded to the population or species level with the use of Hierarchical Generalized Additive Models (HGAMs) with individual-level intercepts and smooth terms [see models "GS" and "GI" in @pedersen_hierarchical_2019]. Such models would allow one to test whether members of the same population or species respond similarly to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior. Individuals and populations that were once thought to have different spatial needs due to differences in behavior may instead be simply responding to different local conditions along the functions in figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) [e.g., @singh_migration_2012]. In the following sections, we illustrate how one can apply this framework to empirical data.

We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs. Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$. Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals' spatial needs under different scenarios.

@@

\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e. its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary. `mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs. Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families (see the help on `?mgcv::family.mgcv` and `?mgcv::twlss` for more information).

The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative. However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g. available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and bounded ratios [e.g., NDVI, see @pettorelli_normalized_2011].

The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass<!-- (e.g., kg of carbon per m$^2$)-->, or energetic intake<!-- (e.g., kcal/day)-->. The Tweedie location-scale family is similar to the Gamma family, but it allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density [but see zero-inflated distributions: @zuur_zero-truncated_2009]. In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family (not available in `mgcv` at the time of publication). If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].

Modeling the mean and variance terms of $R$ should be done carefully. Since trends in both $\e(R)$ and $\var(R)$ can be spatiotemporally nonlinear and non-monotonic, we suggest using a GAM rather than a GLM. However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the mean's terms. An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated. Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt is a reasonable starting point. Additionally, setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms and using restricted marginal likelihood [@wood_fast_2011] should provide reasonably accurate results. We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals. Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model. @simpson_modelling_2018 provides a useful introduction to GAMs for biological time series.

@@

\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the **linear transformation**

@@

\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data]. <!--Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

@@

\noindent As discussed in the previous section, the first step to modeling $H$ is to choose an appropriate family of distributions. Since $H$ is strictly positive and continuous, a Gamma family seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017], but the Tweedie family would also be appropriate if the variance does not scale with the squared mean.

As with the simulated effects of $\mu(t)$ and $\sigma^2(t)$ on $H$ (figure \@ref(fig:5-5-reg)), a location-scale Gamma model is likely required. If there is not sufficient data to fit a location-scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best family of distributions using an information criterion such as the Akaike Information Criterion [@akaike_new_1974]. Due to potential correlation between $\e(R)$ and $\var(R)$ in a small and autocorrelated sample, the relationship between $R$ and $H$ should be modeled carefully and with parsimony. To appropriately constrain the shape of the effects of $\e(R)$ and $\var(R)$, we suggest using a GLM or a shape-constrained GAM (as long as the wiggliness of each term is chosen carefully). Here, we model the effects of $\e(R)$ and $\var(R)$ on $H$ using a shape-constrained GAM with via the `scam` package [@pya_scam_2022]. Additional information on the model can be found in Appendix 2.

@@

```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance. (a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir. (b.) Varince in resource abundance, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir. (c.) Estimated seven-day home range based on the 95\\% utilization quantiles. (d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size. The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them. (f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated. Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$. See Appendix 2 for additional information. The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
knitr::include_graphics('../figures/tapir-example.png')
```

\noindent Figure \@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$. Panels **a** and **b** show the changes in mean and variance in NDVI, respectively, during the tracking period. Panel **c** shows the changes in the tapir's space use over time. Note how the home-range size is larger during periods of lower NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018). In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (January 2018). Finally, panels **d** and **e** show the estimated (partial) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use. Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)). Thus, while the time series shown in panel **c** may appear complicated, the tapir's response can be well described by two simple functions.

@@

<!-- \noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs. The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as new findings. -->

@@

\noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events that occur at the scale of the organism [i.e. approximately one or more times per generation and on the spatial scale of the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [@foley_severe_2008; @berger_climate_2018; @geremia_migrating_2019]. Consequently, organisms with long generations (e.g., elephants, *Loxodonta sp.*) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus are also more likely to respond to and prepare for them. Animals may adapt by adjusting the timing of reproduction [@southwood_habitat_1977; @boersma_wildfire_2021] and torpor or hibernation [@inouye_climate_2000; @goldberg_hibernation_2021], or storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [@armitage_energetics_2003; @nespolo_why_2022] or resource caching [@post_caching_2006]. Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory [@fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; @geremia_migrating_2019; @abrahms_memory_2019; @cavedon_genomic_2022; @foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; @ranc_memory_2022], which is required for migration [@mueller_social_2013; @geremia_migrating_2019; @merkle_spatial_2019; @abrahms_memory_2019; @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
<!-- 
with species names:
